Neighborhoods and Felony Disenfranchisement: The Case of New York City

### 2019-07-24 markdown/felony_disenfranchisement_nys.Rmd
write

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
header-includes:
    - \usepackage{setspace}\doublespacing
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->
\pagenumbering{gobble}
\tableofcontents
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
# Introduction {-}

The political history of the United States has been characterized by a general, if unlinear, trend toward universal sufferage [see, for instance, @Keyssar2009]. At the time of the nation's founding, access to the ballot box was restricted to landed White men; over the following two centuries, the franchise was greatly expanded. Today, voting rights are considered foundational aspects of full citizenship in a liberal democracy. And yet, despite the United State's purported march toward ever-more-inclusive systems of democracy, one large group of American citizens is formally barred from voting. In most of the United States, citizens convicted of felonies are at least temporarily prohibited from casting ballots in elections [@bcj_laws]. Although some states such as Florida and Louisiana have gradually moved to dismantle their systems of felony disenfranchisement, an estimated `r 6.1 - 1.4` million American citizens remain disenfranchised today [@sentencing_2016]. These disenfranchising policies provide a stark reminder that our democracy is not as inclusive as we might like to tell ourselves.

The disenfranchisement of citizens convicted of felony offenses is particularly troubling given the racialized and place-based patterns of policing and incarceration in the United States. As Michelle Alexander [-@Alexander2012] and others have shown, mass incarceration in the post-Civil Rights era has been a tool used by the state to exert control over minority --- and particularly Black --- Americans. This is clear in the case of states such as New York: according to data from the New York State Department of Corrections and Community Supervision, `r round(readRDS("./temp/share_black_incarcerated.rds"), digits = 3) * 100` percent of individuals who were incarcerated in December of 2018 were Black, although the Census Bureau estimates that just 14.6 percent of the citizen voting age population in the state is Black. This disparity is likely due *not* to any inherent differential propensity to commit crimes among different racial groups, but rather to systems of policing and concentrated poverty. As @Gelman2007 shows, for instance, New York's "stop-and-frisk" policy targeted Black and Latino New Yorkers at rates far higher than Whites, even after controlling for neighborhood variability and race-specific criminal propensity.

@@

Felony disenfranchisement policies are part of an interlocking criminal justice system that disproportionately impacts Black Americans living in certain communities. The effects of disenfranchisement are concentrated in neighborhoods that already suffer from myriad disadvantanges stemming from targeted policing and economic marginalization. The neighborhood-specific implications of felony disenfranchisement, however, largely remain unstudied. A number of studies have explored the effect of imprisonment and disenfranchisement on later political participation [@White2019; @Burch2011; @Gerber2014]. Others have looked at the spillover effects of disenfranchisement on eligible Black voters at the state level [@Bowers2009; @King2016]. With the exception of @Burch2013, however, little attention has been paid to the impact of felony disenfranchisement on political participation at the neighborhood level.

@@

In any discussion of felony disenfranchisement in the United States, it is imperative to acknowledge the central role played by race and white supremacy. As Traci Burch has explained, "If policies restricting the voting rights of offenders disparately affect one racial group or party, it is because such policies were *intended* to" [@Burch2010, p. 4; emphasis in the original]. Previous research has established that the presence of nonwhite potential voters is associated with the implementation of felony disenfranchisement policies and that these policies were often adopted during Jim Crow to reduce the political power of Black Americans [@Behrens2003]. In Florida, for instance, felony disenfranchisement was added to the state constitution in 1868. Afterwards, a lawmaker boasted that the amendment had been added to prevent the state from being "niggerized" [@florida_1868, p. 374].

@@

The primary criminal justice dataset comes from a public records request filed by the author to obtain individual-level incarceration and parole records for individuals sentenced to incarceration in New York State since 1990. The data includes a host of information, including: first, middle, and last name; date of birth; class of offense; incarceration start and end dates; dates of parole; county of committment; and others. This analysis is limited to individuals incarcerated for felony offenses. Individuals convicted of misdemeanors are not disenfranchised in New York State. These data come from the New York State Department of Corrections and Community Supervision (NYSDOCCS). These data are used to determine when individuals were incarcerated or on parole, and when they finished their parole supervision.

@@

## Voter File Data {-}
Most states in the United States are required to maintain files with information on all registered voters. In New York, this information is publicly available from the Board of Elections. It includes information on all registered voters, including: first, middle, and last name; date of birth; home address; vote history; and other information. The New York State Voter File also includes information on voters who were previously registered but have since been purged, either because they moved, died, or were incarcerated for a felony offense. The first section of this paper uses a snapshot of the registered voter file from April 30^th^, 2018. The second section of this paper uses a snapshot from March 3^rd^, 2019.

## Geocoding {-}
Voters' home addresses were converted to latitudes and longitudes using a geocoder provided by SmartyStreets. I then used the statistical software R to map these latitudes and longitudes to census block groups, census tracts, and city council districts using shapefiles publicly available from the Census Bureau and the City of New York.

@@

Here, I explore whether the disenfranchisement of residents with a history of participating in elections is related to neighborhood turnout. In this analysis, "lost voters" are individuals ineligible to cast a ballot on a given election day who have cast a ballot in the previous ten years. Although not all lost voters would have participated if they had not been disenfranchised, past participation is an extremely strong predictor of propensity to vote [@Gerber2003]. As such, these are the individuals most likely to have been impacted not only by incarceration but specifically by felony disenfranchisement. Identifying the effects of felony disenfranchisement in neighborhoods that lost individuals with a record of voting --- individuals who would likely have cast a ballot had they been allowed --- provides insight into whether felony disenfranchisement reduces neighborhood turnout through socialization mechanisms.

@@

These individuals are identified by matching NYSDOCCS data to the voter file. I use first, middle, and last names, and dates of birth, to join these two datasets.^[Matching on first and last names and dates of birth can result in false positive matches, especially in large states like New York. Following @Meredith2013, I discuss the potential impact of these errors in [Appendix A].] Figure \ref{fig:citymap} shows where these lost voters lived before going to prison, with city council districts also included. There were `r comma(readRDS("./temp/lost_count_nyc.rds"))` such lost voters within New York City as of the 2017 general election, and `r comma(nrow(readRDS("./temp/ids_of_lost_voters.rds")))` statewide.

@@

Because the majority of neighborhoods have either zero or one lost voter, I begin testing the effect of lost voters on neighborhood turnout using a matching model. Neighborhoods (defined as census tracts and block groups) are considered treated if they were home to *at least one* lost voter in the 2017 election; they are untreated if no voters were disqualified from the election because of a felony conviction.^[It is worth noting that this definition of lost voter is highly conservative. For instance, some individuals who would have voted may have gone to prison after turning 18 but before their first general election. These "untreated" neighborhoods are likely home to individuals who would have voted if it were not for disenfranchisement policies.] I use a genetic match to match treated to untreated census tracts and block groups [@Sekhon2011], using a series of demographic and political indicators. Estimates of racial characteristics, median income, education, age, population, and share noncitizen^[The Census Bureau does not make noncitizen estimates available at the block group level. As such, block groups are assigned their census tract's share noncitizen for matching purposes.] come from the Census Bureau. Party affiliation rates come from the geocoded voter file. Registration rate is calculated by dividing the number of registered voters (calculated using the voter file) by the voting age population (estimated by the Census Bureau). I include voteshare won by the winning city council representative in 2017 as a proxy for the competitiveness of the local race.^[Where neighborhoods cross council district lines, this measure is the mean competitiveness faced by each voter in the neighborhood] Each treated block group is matched to 30 untreated block groups, and each treated census tract is matched with 10 other tracts. Matching is done with replacement. Tables \ref{tab:blockgroup-table} and \ref{tab:tract-table} present the results of these matches.

@@

Much of the literature has discussed whether felony disenfranchisement is particularly demobilizing for eligible Black voters. I therefore include models which explore any potential difference in treatment effect in neighborhoods where a higher share of the population is Black. Table \ref{tab:match-reg} presents the results of these regression models. Each neighborhood is weighted by its adult population, and robust standard errors are clustered at the level of the match [@Abadie2019].
```{r tract-reg-chunk, echo = F, results = "asis", message = F}
load("./temp/match_reg_block_group.rdata")

@@

stargazer(bg_reg, bg_reg_black, tract_reg, tract_reg_black,
          column.separate = c(2, 2),
          column.labels = c("Block Group Level", "Tract Level"),
          header = F,
          se = list(bg_reg_ses, bg_reg_black_ses, tract_reg_ses, tract_reg_black_ses),
          notes = c("Robust standard errors (clustered at", "match level) in parentheses."),
          type = "latex",
          covariate.labels = c("D(Neighborhood Lost a Voter)", "Share Non-Hispanic Black", "D(Lost Voter) $\\times$ Share Non-Hispanic Black", "Constant"),
          dep.var.labels = c("Turnout Rate"),
          title = "(\\#tab:match-reg) Matching Regression",
          table.placement = "H",
          omit.stat = c("f", "ser"))
```
When neighborhoods are measured at the block group level, a lost voter is negatively associated with turnout in the 2017 election. In block groups with lost voters, turnout was on average `r -1*round(bg_reg[["coefficients"]][["treatTRUE"]]*100, digits = 1)` percentage points lower than in comparable block groups without lost voters. This decrease, however, appears to be entirely concentrated within Black neighborhoods. When the dummy identifying neighborhoods with lost voters is interacted with the share of the neighborhood that is Non-Hispanic Black, the basic treatment dummy becomes nonsignificant. The coefficient on the interaction between treatment and share Black indicates that treated neighborhoods that are largely Black saw turnout that was as much as `r -1*round(bg_reg_black[["coefficients"]][["treatTRUE:nh_black"]]*100, digits = 1)` percentage points lower than similar neighborhoods without lost voters. Considering that the overall turnout rate in block groups with a lost voter was just `r round((sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$v2017) / sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$vap))*100, digits = 1)` percent, this effect is alarmingly high. For every 100 votes cast in a predominantly Black block group with a lost voter, as many as `r round(100 - ((100 * ((sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$v2017) / sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$vap)) + bg_reg_black[["coefficients"]][["treatTRUE:nh_black"]])) / (sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$v2017) / sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$vap))), digits = 1)` votes went uncast. Although these variables are largely nonsignificant at the census tract level, this is not particularly surprising. If lost voters have the largest depressive effect on family members and close neighbors, it makes sense that the effect of losing a voter is locally concentrated.

@@

Matching methodologies, of course, only allow us to test the effect of being treated --- here, losing any voter for the 2017 election. The models above do not allow for different effects on turnout based on *how many* voters a neighborhood lost, potentially understating the impact of felony disenfrancisement in the most hard-hit communities. In Table \ref{tab:trad-reg} below, I adopt a standard ordinary least squares regression to investigate whether lost voters are associated with lower turnout rates in the 2017 election. This regression uses the same covariates used in the matching procedure described above. Each neighborhood is weighted by its voting age population, and robust standard errors are clustered by city council district.^[Where neighborhoods cross city council district lines, they are assigned the district in which most of their voters live for clustering purporses.]

@@

The block groups where these depressive effects are not randomly distributed througout the city. They are highly spatially concentrated in Central Brooklyn, Eastern Queens, and Harlem. Figure \ref{fig:postest} applies the coefficient on $Lost\ Voters \times Share\ Black$ from Model 2 in Table \ref{tab:trad-reg} to the city's block groups. The estimated depressive effect is $-0.017\times Lost\ Voters \times Share\ Black$.

@@

During the 2017 mayoral election, felony disenfranchisement laws were responsible for removing an estimated `r comma(readRDS("./temp/lost_count_nyc.rds"))` from New York City neighborhoods. The spatial concentration of these lost voters is striking: as Tables \ref{tab:blockgroup-table} and \ref{tab:tract-table} make clear, these voters were removed from neighborhoods with significantly lower incomes than the rest of the city. They were also removed from neighborhoods that were far less white than the average neighborhood. The systematic removal of voters from these neighborhoods is troubling. However, more than 900,000 votes were cast in the general election of 2017. Although felony disenfranchisement rules have enormous implications for the individuals targeted by them, it is unlikely that the removal of such a small number of voters --- concentrated though they are --- is likely to have large implications on its own.

As this analysis makes clear, however, felony disenfranchisement reaches beyond the individuals who are incarcerated. Previous literature has established that felony disenfranchisement likely impacts Black turnout at the state level. This analysis demonstrates that these demobilizing effects intersect with geographical space to systematically depress the vote in neighborhoods where voters are being sent to prison. These communities are already home to some of the poorest and most marginalized voters in the city; they are home to residents who have the least ability to influence policy through other means such as large campaign contributions. The fact that felony disenfranchisement policies appear to be disincentivizing participation on election day --- that these policies are further weakening the ability of these neighborhoods to participate in the democratic process --- is cause for alarm.

@@

Our variable of interest, therefore, is likely correlated with our regression error. Such a correlation inhibits our ability to simply regress turnout on a dummy indicating whether an individual's rights were restored. The standard way of dealing with such a problem is an instrumental variables approach. Such an approach has been widely used in the context of voter turnout [e.g. @Gerber2000; @Green2012]. A valid instrumental variable must satisfy two criteria: it must be correlated with the right-hand-side (endogenous) variable of interest, and it must be uncorrelated with the error of the regression.

@@

\newpage
# Appendix A {-}

>>>>

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{rotating}
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->
\pagenumbering{gobble}
\tableofcontents
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
# Introduction {-}

The political history of the United States has been characterized by a general, if unlinear, trend toward universal sufferage [see, for instance, @Keyssar2009]. At the time of the nation's founding, access to the ballot box was restricted to landed White men; over the following two centuries, the franchise was greatly expanded. Today, voting rights are considered foundational aspects of full citizenship in a liberal democracy. Today, despite the United State's march toward ever-more-inclusive systems of democracy, just one large group of American citizens is formally barred from voting. In most of the United States, citizens convicted of felonies are at least temporarily prohibited from casting ballots in elections [@bcj_laws]. Although some states such as Florida and Louisiana have gradually moved to dismantle their systems of felony disenfranchisement, an estimated `r 6.1 - 1.4` million American citizens remain disenfranchised today [@sentencing_2016].

The disenfranchisement of citizens convicted of felony offenses is intersects with the racialized and place-based patterns of policing and incarceration in the United States. As Michelle Alexander [-@Alexander2012] and others have shown, mass incarceration in the post-Civil Rights era has been used to exert control over minority --- and particularly Black --- Americans. In states such as New York, for instance, the overrepresentation of minorities among the incarcerated is striking: according to data from the New York State Department of Corrections and Community Supervision, `r round(readRDS("./temp/share_black_incarcerated.rds"), digits = 3) * 100` percent of individuals who were incarcerated in December of 2018 were non-Hispanic Black, although the Census Bureau estimates that just 14.3 percent of the citizen voting age population in the state is non-Hispanic Black.^[Latinos are also overrepresented among the incarcerated population, though not as dramatically: Latinos make up 14.1 percent of the citizen voting age population and 22.9 percent of the incarcerated population.] This disparity is likely due not to any inherent differential propensity to commit crimes among different racial groups, but rather to systems of policing and concentrated poverty. As @Gelman2007 shows, for instance, New York's "stop-and-frisk" policy impacted Black and Latino New Yorkers at rates far higher than Whites, even after controlling for neighborhood variability and race-specific criminal propensity.

@@

Felony disenfranchisement policies are part of an interlocking criminal justice system that disproportionately impacts Black Americans living in certain communities. The effects of disenfranchisement are concentrated in neighborhoods that already suffer from myriad disadvantanges thanks to social and economic marginalization. The neighborhood-specific implications of felony disenfranchisement, however, largely remain unstudied. A number of studies have explored the effect of imprisonment and disenfranchisement on later political participation [@White2019; @Burch2011; @Gerber2014]. Others have looked at the spillover effects of disenfranchisement on eligible Black voters at the state level [@Bowers2009; @King2016]. With the exception of @Burch2013, however, little attention has been paid to the impact of felony disenfranchisement on political participation at the neighborhood level. Filling this gap in the literature is of great importance.

@@

In any discussion of felony disenfranchisement in the United States, it is important to acknowledge the central role played by race. As Traci Burch has explained, "If policies restricting the voting rights of offenders disparately affect one racial group or party, it is because such policies were *intended* to" [@Burch2010, p. 4; emphasis in the original]. Previous research has established that the presence of nonwhite potential voters is associated with the implementation of felony disenfranchisement policies and that these policies were often adopted during Jim Crow to reduce the political power of Black Americans [@Behrens2003]. In Florida, for instance, felony disenfranchisement was added to the state constitution in 1868. Afterwards, a lawmaker boasted that the amendment had been added to prevent the state from being "niggerized" [@florida_1868, p. 374].

@@

The primary criminal justice dataset comes from a public records request filed by the author to obtain individual-level incarceration and parole records for individuals who have been incarcerated in New York State since 1990. The data includes a host of information, including: first, middle, and last name; date of birth; class of offense; incarceration start and end dates; dates of parole; county of committment; and others. These data come from the New York State Department of Corrections and Community Supervision (NYSDOCCS) and are used to determine when individuals were incarcerated or on parole, and when they finished their parole supervision.

The state makes records available only for individuals who have been incarcerated for felony offenses. It does not make information about individuals sentenced to probation or incarcerated for misdemeanors. Thus, while the data covers all individuals subject to felony disenfranchisement rules (only individuals incarcerated for felony offenses lose their voting rights), it limits the availability of a potentially helpful control group. It also only includes individuals who are held in state prisons.

@@

## Voter File Data {-}
Most states in the United States are required to maintain files with information on all registered voters. In New York, this information is publicly available from the Board of Elections. It includes information on all registered voters, including: first, middle, and last name; date of birth; home address; vote history; and other information. The first section of this paper uses a snapshot of the registered voter file from April 30^th^, 2018. The second section of this paper uses a snapshot from March 3^rd^, 2019.

The New York State Voter File is unique in its treatment of "purged" voters: although most states remove voters from their voter files once they are no longer eligible to vote, New York continues to include them in the file (but marks them as purged). I can therefore identify voters who were registered in the past but have since been purged due to a felony conviction.

## Geocoding {-}
Voters' home addresses were converted to latitudes and longitudes using a geocoder provided by SmartyStreets. I then used the statistical software R to map these latitudes and longitudes to census block groups, census tracts, and city council districts using shapefiles publicly available from the Census Bureau and the City of New York. This geocoder is not perfect: among individuals registered to vote in New York City, the geocoder failed to determine the latitude and longitude of the addresses of `r round(readRDS("./temp/bad_geocode_nyc.rds") * 100, digits = 1)` percent of registered voters. The geocoder was slightly less successful when it came to lost voters (defined and discussed below); `r round(readRDS("./temp/bad_geo_lost_voters.rds") * 100, digits = 1)` percent of these individuals were not geocoded. Voters who were not successfully geocoded are dropped from the analysis; however, because so few observations went uncoded, it is unlikely to affect the analysis.

## Matching {-}
Registered and formerly registered individuals who have been to prison in New York are identified by matching the NYSDOCCS records with the registered voter file. I match individuals in each dataset using first name, middle name, last name, and date of birth. To be considered a "match," records must have the exact same birth date. The first and last names must also be exact matches (conditional on the adjustments discussed below). The middle names must meet one of the following conditions in order to qualify:

* Middle names are identical. If neither set of records includes a middle name, this condition is met.
* A full middle name in one set of records and only a middle initial in the other. The first letter of the full middle name must be the same as the middle initial in the other set of records.
* A middle name or middle initial in one set of records, and a missing middle name in the other set.

Thus, "John Andrew Doe" and "John A Doe" would count as matches. Similarly, "John Andrew Doe" and "John Doe" would count, while "John Andew Doe" and "John Anthony Doe" would not.

There are two types of potential error in this methdology: a false positive will result when a parolee's records matches the record of a voter who is a different individual but shares the same name and date of birth. False negatives will occur when an individual has a different name in the different sets of records, or when the birthdate is incorrectly reported in one of the sets of records

Testing for the presence of false positive matches is fairly straightforward. @Meredith2013 offers one way to test their prevalence using placebo matching. I slightly alter the date of birth reported in the parole discharge dataset to create false records. Comparing the number of matches between these "fake" discharge records and the voter file with the number of matches between the "true" records and the voter file provides an estimate of how frequently false positives occur. Table \ref{tab:change-dobs} shows the results of true matches, as well as when I construct a set of fake records by adding or subtracting 35 days from a parolee's birthdate. This analysis indicates that false positives account for between 0.6 and 0.7 percent of all matches, a share that is likely too small to have any material impact on the overall analysis.

```
Testing for false negatives is more challenging. If an individual marries and changes her name after being discharged from parole, for instance, I will not identify her using my matching methodology. Similarly, "John Doe" and "Jonathan Doe" would not result in a match. To reduce the likelihood of these false negatives I remove all punctuation from all names, and standardize capitalization. A record with a last name of "O'Donnell" in one dataset, therefore, would match a last name of "O DONNELL" in the other (provided the other criteria are satisfied). Such standardizations, however, will miss individuals who change their names entirely. For three reasons, however, this is not likely to present major challenges: firstly, women are far more likely to change their last names than men, and women make up barely 6 percent of individuals who have been discharged from felony parole. Secondly, because both parolee discharge and voter registration are legal records, individuals are likely to be recorded using their full names (that is to say, an individual is unlikely to be "John" in one set of records and "Jonathan" in the other). Finally, rates of false negatives are likely to be constant within the state during the study period, and there is no reason to believe that these false negatives would be associated with being discharged from parole after the Executive Order went into effect.

@@

Here, I explore whether the disenfranchisement of residents with a history of participating in elections is related to neighborhood turnout. In this analysis, "lost voters" are individuals ineligible to cast a ballot on a given election day who have cast a ballot in the previous ten years. Although not all lost voters would have participated if they had not been disenfranchised, past participation is an extremely strong predictor of propensity to vote [@Gerber2003]. As such, these are the individuals most likely to have been impacted not only by incarceration but specifically by felony disenfranchisement. Identifying the effects of felony disenfranchisement in neighborhoods that lost individuals with a record of voting --- individuals who would likely have cast a ballot had they been allowed --- provides insight into whether felony disenfranchisement reduces neighborhood turnout through socialization mechanisms. Because New York State's voter file includes information on individuals who have been purged for felony convictions, I can reconstruct the vote history even for voters who are no longer eligible to vote.

@@

Figure \ref{fig:citymap} shows where these lost voters lived before going to prison, with city council districts also included. There were `r comma(readRDS("./temp/lost_count_nyc.rds"))` such lost voters within New York City as of the 2017 general election, and `r comma(nrow(readRDS("./temp/ids_of_lost_voters.rds")))` statewide.

@@

Because the majority of neighborhoods have either zero or one lost voter, I begin testing the effect of lost voters on neighborhood turnout using a matching model. Neighborhoods (defined as census tracts and block groups) are considered treated if they were home to *at least one* lost voter in the 2017 election; they are untreated if no voters were disqualified from the election because of a felony conviction.^[It is worth noting that this definition of lost voter is highly conservative. For instance, some individuals who would have voted may have gone to prison after turning 18 but before their first general election. These "untreated" neighborhoods are likely home to individuals who would have voted if it were not for disenfranchisement policies.] I use a genetic match to match treated to untreated census tracts and block groups [@Sekhon2011], using a series of demographic and political indicators. Estimates of racial characteristics, median income, education, age, population, and share noncitizen^[The Census Bureau does not make noncitizen estimates available at the block group level. As such, block groups are assigned their census tract's share noncitizen for matching purposes.] come from the Census Bureau. Party affiliation rates come from the geocoded voter file. Registration rate is calculated by dividing the number of registered voters (calculated using the voter file) by the voting age population (estimated by the Census Bureau). I include voteshare won by the winning city council representative in 2017 as a proxy for the competitiveness of the local race;^[Where neighborhoods cross council district lines, this measure is the mean competitiveness faced by each voter in the neighborhood] where city council district races were more competitive, I expect that more voters will have turned out. Each treated block group is matched to 30 untreated block groups, and each treated census tract is matched with 10 other tracts. Matching is done with replacement. Tables \ref{tab:blockgroup-table} and \ref{tab:tract-table} present the results of these matches.

@@

Much of the literature has discussed whether felony disenfranchisement is particularly demobilizing for eligible Black voters. I therefore include models which explore any potential difference in treatment effect in neighborhoods where a higher share of the population is Black. Table \ref{tab:match-reg} presents the results of these regression models. Robust standard errors are clustered at the level of the match [@Abadie2019].^[The observations in these models are not weighted by voting age population. Such weighting materially impacts neither the size or statistical significance of the coefficients of interest.]
```{r tract-reg-chunk, echo = F, results = "asis", message = F}
load("./temp/match_reg_block_group.rdata")

@@

stargazer(bg_reg, bg_reg_black, tract_reg, tract_reg_black, tract_reg_black,
          column.separate = c(2, 2),
          column.labels = c("Block Group Level", "Tract Level"),
          header = F,
          se = list(bg_reg_ses, bg_reg_black_ses, tract_reg_ses, tract_reg_black_ses),
          notes = c("Robust standard errors (clustered at", "match level) in parentheses."),
          type = "latex",
          covariate.labels = c("D(Neighborhood Lost a Voter)", "Share Non-Hispanic Black", "D(Lost Voter) $\\times$ Share Non-Hispanic Black", "Constant"),
          dep.var.labels = c("Turnout Rate"),
          title = "(\\#tab:match-reg) Matching Regression",
          table.placement = "H",
          omit.stat = c("f", "ser"))
```
When neighborhoods are measured at the block group level, a lost voter is negatively associated with turnout in the 2017 election. In block groups with lost voters, turnout was on average `r -1*round(bg_reg[["coefficients"]][["treatTRUE"]]*100, digits = 1)` percentage points lower than in comparable block groups without lost voters. This decrease, however, appears to be entirely concentrated within Black neighborhoods. When the dummy identifying neighborhoods with lost voters is interacted with the share of the neighborhood that is Non-Hispanic Black, the basic treatment dummy becomes nonsignificant. The coefficient on the interaction between treatment and share Black indicates that treated neighborhoods that are largely Black saw turnout that was as much as `r -1*round(bg_reg_black[["coefficients"]][["treatTRUE:nh_black"]]*100, digits = 1)` percentage points lower than similar neighborhoods without lost voters. Considering that the overall turnout rate in block groups with a lost voter was just `r round((sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$v2017) / sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$vap))*100, digits = 1)` percent, this effect is alarmingly high. For every 100 votes cast in an entirely Black block group with a lost voter, as many as `r round(100 - ((100 * ((sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$v2017) / sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$vap)) + bg_reg_black[["coefficients"]][["treatTRUE:nh_black"]])) / (sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$v2017) / sum(filter(readRDS("./temp/block_group_pre_match.rds"), treat == T)$vap))), digits = 1)` votes went uncast. The effects are slightly larger when neighborhoods are measured as block groups than as census tracts. This is not surprising --- because the spillover effects are likely to operate through social networks, smaller geographical units are likely to be more effected.

@@

Matching methodologies, of course, only allow us to test the effect of being treated --- here, losing any voter for the 2017 election. The models above do not allow for different effects on turnout based on *how many* voters a neighborhood lost, potentially understating the impact of felony disenfrancisement in the most hard-hit communities. In Table \ref{tab:trad-reg} below, I adopt a standard ordinary least squares regression to investigate whether lost voters are associated with lower turnout rates in the 2017 election. This regression uses the same covariates used in the matching procedure described above. Robust standard errors are clustered by city council district.^[Where neighborhoods cross city council district lines, they are assigned the district in which most of their voters live for clustering purporses. The observations in these models are not weighted by voting age population. Such weighting materially impacts neither the size or statistical significance of the coefficients of interest.]

@@

The block groups where these depressive effects are not randomly distributed througout the city. They are highly spatially concentrated in Central Brooklyn, Eastern Queens, and Harlem. Figure \ref{fig:postest} applies the coefficient on $Lost\ Voters \times Share\ Black$ from Model 2 in Table \ref{tab:trad-reg} to the city's block groups. The estimated depressive effect is $-0.018\times Lost\ Voters \times Share\ Black$.

@@

During the 2017 mayoral election, felony disenfranchisement laws were responsible for removing an estimated `r comma(readRDS("./temp/lost_count_nyc.rds"))` from New York City neighborhoods. The spatial concentration of these lost voters is striking, as demonstrated in \ref{fig:citywide-map}. The systematic removal of voters from these neighborhoods is troubling. However, more than one million votes were cast in the general election of 2017. Although felony disenfranchisement rules have enormous implications for the individuals targeted by them, the removal of such a small number of voters --- concentrated though they are --- is unlikely to have large implications on its own.

As this analysis makes clear, however, felony disenfranchisement reaches beyond the individuals who are incarcerated. Previous literature has established that felony disenfranchisement likely impacts Black turnout at the state level. This analysis demonstrates that these demobilizing effects intersect with geographical space to systematically depress the vote in neighborhoods where voters are being sent to prison. As discussed above, these neighborhoods have far lower incomes than the rest of the city. The median income of a block groups without lost voters is more than 40 percent higher than the median income of block groups with lost voters. Similarly, just 17 percent of individuals in the average block group with a lost voter were white, compared with 40 percent of the population in other block groups. Felony disenfranchisement has spillover effects on neighborhood turnout, and these neighborhoods systematically differ from the rest of the city's population. The effects are concentrated in neighborhoods where the most marginalized members of society live. These highly concentrated spillover effects are cause for major concern.

Of equal concern is the apparent concentration of these effects in Black neighborhoods. Both Tables \ref{tab:match-reg} and \ref{tab:trad-reg} indicate that once the lost voter indicator is interacted with the share of a neighborhood that is Non-Hispanic Black, the lost voter indicator is either insignificant or significant and positive. This means that lost voters have no spillover effects in neighborhoods with small Black populations. Understanding the magnitude of this finding is important. In New York City, there are `r readRDS("./temp/count_gt_90_black")` block groups where the Black community makes up more than 90 percent of the population. On average, the voting age population in these neighborhoods was `r round(readRDS("./temp/vap_gt_90_black"))`. Model 2 of Table \ref{tab:match-reg} indicates that, in a block group that is 90 percent Black with a voting-age population of `r round(readRDS("./temp/vap_gt_90_black"))`, having any lost voter reduced turnout by `r round(0.9 * -1*bg_reg_black[["coefficients"]][["treatTRUE:nh_black"]] * readRDS("./temp/vap_gt_90_black"), digits = 1)` ballots. Conversely, in a block group with the same voting age population that is just 10 percent Black, the loss of any voter decreased turnout by just `r round(0.1 * -1*bg_reg_black[["coefficients"]][["treatTRUE:nh_black"]] * readRDS("./temp/vap_gt_90_black"), digits = 1)` ballots.

The standard OLS regression results tell a similar story: in our average block group that is 90 percent Black, each lost voter cost the neighborhood `r round(0.9 * -1*bg_model2[["coefficients"]][["lost_voters_black"]] * readRDS("./temp/vap_gt_90_black"), digits = 1)` ballots. In block groups of the same size where the population is just 10 percent Black, lost voters cost the neighborhood just `r round(0.1 * -1*bg_model2[["coefficients"]][["lost_voters_black"]] * readRDS("./temp/vap_gt_90_black"), digits = 1)` ballots --- hardly more votes than the lost voter herself.

Why do lost voters in Black neighborhoods have such large spillover effects, when lost voters in predominantly non-Black neighborhoods do not? Much of the previous literature in this space establishes that individuals who have negative interactions with the government are less likely to choose to interact with the state in the future. @Lerman2013, for instance, shows that neighborhoods where there are many police stops that involve searches or use of force use 311 services less frequently. @Weaver2010 argues that interactions with the criminal justice system changes how individuals understand both their governments and their identities as citizens. It is possible that Black communities understand the loss of voters differently than non-Black communities, and therefore respond differently.

Individuals who live in neighborhoods where police activity is relatively limited may interpret the incarceration of a neighbor as a largely individual phenomenon. If it is understood as an isolated or individual event, voters in these neighborhoods who are not incarcerated are not likely to update their view of the state. They may draw no connections between their neighbor's imprisonment and their own efficacy as a voter.

In the neighborhoods where policing is most prevalent --- often, lower-income Black communities --- the incarceration of a neighbor might not be interepreted so individualistically. It may, rather, be interpreted as another reminder of the government's unfairness. If a would-be voter finds herself soured on political participation because of her neighbor's incarceration, she may be less likely to cast a ballot.

Interestingly, this finding mirrors @White2019, which finds that brief jail spells decrease future participation more for Black individuals than for White individuals. This is perhaps unsurprising. A large body of research indicates that, even after controlling for various sociodemographic characteristics and interactions with the police, Black Americans have far more negative views of the criminal justice system than White Americans [e.g. @Browning1992; @Henderson1997; @Wu2009]. It should come as no surprise that experience with the criminal justice system less likely to be viewed in isolation (and therefore more likely to effect voting patterns) in Black neighborhoods.

Whether this finding is surprising or not, it is troubling.

@@

Our variable of interest, therefore, is likely correlated with our regression error. Such a correlation inhibits our ability to simply regress turnout on a dummy indicating whether an individual's rights were restored. The standard way of dealing with such a problem is an instrumental variables approach. Such an approach has been widely used in the context of voter turnout [e.g. @Ansolabehere1999; @Gerber2000; @Milligan2004]. A valid instrumental variable must satisfy two criteria: it must be correlated with the right-hand-side (endogenous) variable of interest, and it must be uncorrelated with the error of the regression.

@@
